---
title: "Package Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette covers the entire adversarial random forest (ARF) pipeline, from model training to parameter learning, density estimation, and data synthesis.

# Adversarial Training

The ARF algorithm is an iterative procedure. In the first instance, we generate synthetic data by independently sampling from the marginals of each feature and training a random forest (RF) to distinguish original from synthetic samples. If accuracy is greater than $0.5 + \delta$ (where `delta` is a user-controlled tolerance parameter, generally set to 0), we create a new dataset by sampling from the marginals within each leaf and training another RF classifier. The procedure repeats until original and synthetic samples cannot be reliably distinguished. With the default `verbose = TRUE`, the algorithm will print accuracy at each iteration. 

```{r arf}
# Load libraries
library(arf)
library(data.table)
library(ggplot2)

# Set seed
set.seed(123, "L'Ecuyer-CMRG")

# Train ARF
arf <- adversarial_rf(iris)
```

The printouts can be turned off by setting `verbose = FALSE`. Accuracy is still stored within the `arf` object, so you can evaluate convergence after the fact. The warning appears just once per session. It can be suppressed by setting `parallel = FALSE` or registering a parallel backend (more on this below).

```{r arf2, fig.height=5, fig.width=5}
# Train ARF with no printouts
arf <- adversarial_rf(iris, verbose = FALSE)

# Plot accuracy against iterations (model converges when accuracy <= 0.5)
tmp <- data.frame('acc' = arf$acc, 'iter' = seq_len(length(arf$acc)))
ggplot(tmp, aes(iter, acc)) + 
  geom_point() + 
  geom_path() +
  geom_hline(yintercept = 0.5, linetype = 'dashed', color = 'red') 
```

We find a quick drop in accuracy following the resampling procedure, as desired. If the ARF has converged, then resulting splits should form fully factorized leaves, i.e. subregions of the feature space where variables are locally independent. 

ARF convergence is almost surely guaranteed as $n \rightarrow \infty$ (see [Watson et al., 2022](https://arxiv.org/abs/2205.09435), Thm. 1). However, this has no implications for finite sample performance. In practice, we often find that adversarial training completes in just one or two rounds, but this may not hold for some datasets. To avoid infinite loops, users can increase the slack parameter `delta` or set the `max_iters` argument (default = 10). In addition to these failsafes, `adversarial_rf` uses early stopping by default `(early_stop = TRUE)`, which terminates training if factorization does not improve from one round to the next. This is recommended, since discriminator accuracy rarely falls much lower once it has increased. 

For density estimation tasks, we recommend increasing the default number of trees. We generally use 100 in our experiments, though this may be suboptimal for some datasets. Likelihood estimates are not very sensitive to this parameter above a certain threshold, but larger models incur extra costs in time and memory. We can speed up computations by registering a parallel backend, in which case ARF training is distributed across cores using the `ranger` package. Much like with `ranger`, the default behavior of `adversarial_rf` is to compute in parallel if possible. How exactly this is done varies across operating systems. The following code works on Unix machines.

```{r par, eval=FALSE}
# Register cores - Unix
library(doParallel)
registerDoParallel(cores = 2)
```

Windows requires a different setup.

```{r par2, eval=FALSE}
# Register cores - Windows
library(doParallel)
cl <- makeCluster(2)
registerDoParallel(cl)
```

In either case, we can now execute in parallel.

```{r arf3}
# Rerun ARF, now in parallel and with more trees
arf <- adversarial_rf(iris, num_trees = 100)
```

The result is an object of class `ranger`, which we can input to downstream functions. 

# Parameter Learning

The next step is to learn the leaf and distribution parameters using forests for density estimation (FORDE). This function calculates the coverage, bounds, and pdf/pmf parameters for every variable in every leaf. This can be an expensive computation for large datasets, as it requires $\mathcal{O}\big(B \cdot d \cdot n \cdot \log(n)\big)$ operations, where $B$ is the number of trees, $d$ is the data dimensionality, and $n$ is the sample size. Once again, the process is parallelized by default. 

```{r forde}
# Compute leaf and distribution parameters
params <- forde(arf, iris)
```

Default behavior is to use a truncated normal distribution for continuous data (with boundaries given by the tree's split parameters) and a multinomial distribution for categorical data. We find that this produces stable results in a wide range of settings. You can also use a uniform distribution for continuous features by setting `family = 'unif'`, thereby instantiating a piecewise constant density estimator. 

```{r forde_unif}
# Recompute with uniform density
params_unif <- forde(arf, iris, family = 'unif')
```

This method tends to perform poorly in practice, and we do not recommend it. The option is implemented primarily for benchmarking purposes. Alternative families, e.g. truncated Poisson or beta distributions, may be useful for certain problems. Future releases will expand the range of options for the `family` argument.

The `alpha` and `epsilon` arguments allow for optional regularization of multinomial and uniform distributions, respectively (defaults for both are set to $0.1$). These help prevent zero likelihood samples when test data fall outside the support of training data. The former is a pseudocount parameter that applies Laplace smoothing within leaves, preventing unobserved values from being assigned zero probability unless splits explicitly rule them out. In other words, we impose a flat Dirichlet prior $\text{Dir}(\alpha)$ and report posterior probabilities rather than maximum likelihood estimates. The latter is a slack parameter on empirical bounds that expands the estimated extrema for continuous features by a factor of $1 + \epsilon$.

Compare the results of our original probability estimates for the `Species` variable with those obtained when we set $\alpha = 0$.

```{r dirichlet}
# Recompute without additive smoothing
params_mle <- forde(arf, iris, alpha = 0)

# Compare results
head(params$cat)
head(params_mle$cat)
```

Under Laplace smoothing, extreme probabilities only occur when the splits explicitly demand it. Otherwise, all values shrink toward a uniform prior. Note that these two data tables may not have exactly the same rows, as we omit zero probability events to conserve memory. However, we can verify that probabilities sum to unity for each leaf-variable combination.

```{r unity}
# Sum probabilities
summary(params$cat[, sum(prob), by = .(f_idx, variable)]$V1)
summary(params_mle$cat[, sum(prob), by = .(f_idx, variable)]$V1)
```

The `forde` function outputs a list of length 5, with entries for (1) continuous features; (2) categorical features; (3) leaf parameters; (4) variable metadata; and (5) data input class. 

```{r forde2}
params
```

These parameters can be used for a variety of downstream tasks, such as likelihood estimation and data synthesis.

# Likelihood Estimation

To calculate log-likelihoods, we pass `arf` and `params` on to the `lik` function, along with the data whose likelihood we wish to evaluate.

```{r lik}
# Compute likelihood under truncated normal and uniform distributions
ll <- lik(params, iris, arf = arf)
ll_unif <- lik(params_unif, iris, arf = arf)

# Compare average negative log-likelihood (lower is better)
-mean(ll)
-mean(ll_unif)
```

Note that the piecewise constant estimator does considerably worse in this experiment.

We can compute likelihoods on the probability scale by setting `log = FALSE`, but this may result in numerical underflow. There is also a `batch` argument, which has no impact on results but can be more memory efficient for large datasets. For instance, we could rerun the code above in batches of size 50:

```{r lik2}
# Compute likelihood in batches of 50
ll_50 <- lik(params, iris, arf = arf, batch = 50)

# Identical results?
identical(ll, ll_50)
```

In this example, we have used the same data throughout. This may lead to overfitting. With sufficient data, it is preferable to use a training set for `adversarial_rf`, a validation set for `forde`, and a test set for `lik`. Alternatively, we can set the `oob` argument to `TRUE` for either of the latter two functions, in which case computations are performed only on out-of-bag (OOB) data. These are samples that are randomly excluded from a given tree due to the bootstrapping subroutine of the RF classifier. Note that this only works when the dataset `x` passed to `forde` or `lik` is the same one used to train the `arf`. Recall that a sample's probability of being excluded from a single tree is $e^{-1} \approx 0.368$. When using `oob = TRUE`, be sure to include enough trees so that every observation is likely to be OOB at least a few times. 

# Data synthesis

For this experiment, we use the `smiley` simulation from the `mlbench` package, which allows for easy visual assessment. We draw a training set of $n = 1000$ and simulate $1000$ synthetic datapoints. Resulting data are plotted side by side.

```{r smiley, fig.height=5, fig.width=8}
# Simulate training data
library(mlbench)
x <- mlbench.smiley(1000)
x <- data.frame(x$x, x$classes)
colnames(x) <- c('X', 'Y', 'Class')

# Fit ARF
arf <- adversarial_rf(x, mtry = 2)

# Estimate parameters
params <- forde(arf, x)

# Simulate data
synth <- forge(params, n_synth = 1000)

# Compare structure
str(x)
str(synth)

# Put it all together
x$Data <- 'Original'
synth$Data <- 'Synthetic'
df <- rbind(x, synth)

# Plot results
ggplot(df, aes(X, Y, color = Class, shape = Class)) + 
  geom_point() + 
  facet_wrap(~ Data)
```

The general shape is clearly recognizable, even if some stray samples are evident and borders are not always crisp. This can be improved with more training data.

Note that the default behavior of `adversarial_rf` is to treat integers as ordered factors, with a warning. This makes sense for, say, count data with limited support (e.g., number of petals on a flower). However, this is probably not the desired behavior for other integer variables. Consider the `diamonds` dataset, where `price` is classed as an integer.

```{r price, fig.height=5, fig.width=5}
# Check data
head(diamonds)

# View the distribution
hist(diamonds$price)

# How many unique prices?
length(unique(diamonds$price))
```

This variable should clearly not be treated as a factor with 11602 levels. To make sure we fit a continuous density for price, we re-class the feature as numeric.

```{r price2, fig.height=5, fig.width=5}
# Re-class 
diamonds$price <- as.numeric(diamonds$price)

# Take a random subsample of size 2000
s_idx <- sample(1:nrow(diamonds), 2000)

# Train ARF
arf <- adversarial_rf(diamonds[s_idx, ])

# Estimate parameters
params <- forde(arf, diamonds[s_idx, ])

# Check distributional families
params$meta

# Forge data, check histogram
synth <- forge(params, n_synth = 1000)
hist(synth$price)
```

Using `family = 'truncnorm'`, the distribution for `price` will now be modeled with a truncated Gaussian mixture. Though the general outline of the histogram looks about right, we do find some implausible values, e.g. negative prices. This can be overcome by manually setting a hard lower bound.

```{r lb, fig.height=5, fig.width=5}
# Set price minimum to empirical lower bound
params$cnt[variable == 'price' & min == -Inf, min := min(diamonds$price)]

# Re-forge, check histogram
synth <- forge(params, n_synth = 1000)
hist(synth$price)
```

Alternatively, we could retrain under log transformation.

```{r lprice, fig.height=5, fig.width=5}
# Transform price variable
tmp <- as.data.table(diamonds[s_idx, ])
tmp[, price := log(price)]

# Retrain ARF
arf <- adversarial_rf(tmp)

# Estimate parameters
params <- forde(arf, tmp)

# Forge, check histogram
synth <- forge(params, n_synth = 1000)
hist(exp(synth$price))
```

This may be unnecessary with sufficiently large sample sizes. For instance, negative prices are exceedingly rare when training on the complete diamonds dataset ($n = 53940$). However, if natural constraints are known *a priori*, they can easily be incorporated into `params`. 

