---
title: "arf benchmarks"
subtitle: "Version comparison, backend comparison, and the release trend"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    embed-resources: true
    fig-width: 11
    fig-height: 7
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
suppressMessages({
  library(data.table)
  library(ggplot2)
})
source("run-cell.R")
source("collate.R")

# Small tables render as plain HTML; big ones get a sortable, filterable widget
# rather than a wall of rows.
bench_tbl <- function(x, big = 25L, page = 15L) {
  x <- as.data.frame(x)
  if (!nrow(x)) {
    return(htmltools::HTML("<p><em>No rows.</em></p>"))
  }
  if (nrow(x) >= big && requireNamespace("reactable", quietly = TRUE)) {
    reactable::reactable(
      x,
      filterable = TRUE, sortable = TRUE, searchable = TRUE,
      defaultPageSize = page, compact = TRUE, striped = TRUE, highlight = TRUE,
      defaultColDef = reactable::colDef(minWidth = 70)
    )
  } else {
    htmltools::HTML(knitr::kable(x, format = "html", table.attr = 'class="table table-sm"'))
  }
}

bench_theme <- function(base_size = 15) {
  theme_minimal(base_size = base_size) +
    theme(
      legend.position = "bottom",
      legend.title = element_blank(),
      plot.title.position = "plot",
      # Facet labels on the left rather than the right, where they read as
      # detached from the panel they describe.
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0),
      panel.spacing = unit(1, "lines")
    )
}

# Newest version-comparison CSV, or one named via QUARTO_BENCH_CSV.
bench_csv <- Sys.getenv("QUARTO_BENCH_CSV")
if (!nzchar(bench_csv)) {
  cands <- list.files("results", pattern = "^bench-.*\\.csv$", full.names = TRUE)
  bench_csv <- if (length(cands)) cands[order(basename(cands), decreasing = TRUE)][1] else ""
}
run <- if (nzchar(bench_csv) && file.exists(bench_csv)) as.data.table(fread(bench_csv)) else NULL
baseline <- Sys.getenv("QUARTO_BENCH_BASELINE", "main")
have_run <- !is.null(run) && baseline %in% run$ref
```

::: {.callout-note collapse="true"}
## What the memory numbers mean

Every cell is measured twice per round: once running the operation, and once
running nothing but the setup (an R interpreter, `arf`, and the fixture data).
That second measurement is the **floor**, about 200 MB here. The **marginal**
is peak minus floor: what the operation itself allocated, as opposed to the
fixed cost of having R and the package loaded.

Deltas are computed on the marginal, because the floor is identical across refs
and would dilute any real difference several-fold. A marginal smaller than its
own floor is below what this instrument can resolve, and gets no verdict.
:::

## Latest version comparison

```{r}
#| label: latest-run
if (is.null(run)) {
  cat("No `results/bench-*.csv` yet. Run `bench/arf-bench run`.\n")
} else if (!have_run) {
  cat("Baseline `", baseline, "` is not in ", basename(bench_csv),
    "; set `--base` or QUARTO_BENCH_BASELINE.\n", sep = "")
} else {
  cat(sprintf(
    "`%s` | %d rows | refs: %s | metric: %s | host(s): %s\n\n",
    basename(bench_csv), nrow(run), paste(unique(run$ref), collapse = ", "),
    paste(unique(run$metric), collapse = ", "),
    paste(unique(run$host), collapse = ", ")
  ))
}
```

```{r}
#| label: latest-table
if (have_run) {
  d <- as.data.table(bench_deltas(run, baseline = baseline))[ref != baseline]
  bench_tbl(d[order(op, n, trees, backend, workers, ref), .(
    op, backend, w = workers, n, trees, ref,
    marginal_MB = round(peak_delta_mb, 1),
    floor_MB = round(floor_mb, 1),
    mem_pct = round(delta_mem_pct, 1),
    mem = mem_verdict,
    time_s = round(time_median, 3),
    time_pct = round(delta_time_pct, 1),
    time = time_verdict
  )])
}
```

## Backend comparison

Which backend, at which worker count. This is a live question, not a historical
one: `foreach` forks and copies, `mirai` shares read-only through `mori`, and
which wins depends on data size and worker count. Worker counts are discrete
measurements, so the connecting lines are dotted: they group a series, they do
not interpolate.

```{r}
#| label: backend-compare
#| output: asis
#| fig-height: 9
if (!have_run) {
  cat("No run loaded.\n")
} else if (uniqueN(run$backend) < 2L) {
  cat("This run used a single backend (`", unique(run$backend),
    "`), so there is nothing to compare. The full tier varies backend and workers.\n", sep = "")
} else {
  par_run <- run[backend != "sequential" & !is.na(workers)]
  seq_run <- run[backend == "sequential"]
  long <- rbind(
    par_run[!is.na(peak_delta_mb), .(op, n, trees, backend, workers, ref,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    par_run[!is.na(time_median), .(op, n, trees, backend, workers, ref,
      metric = "time (s)", value = time_median)]
  )
  seq_long <- rbind(
    seq_run[!is.na(peak_delta_mb), .(op, n, trees, ref,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    seq_run[!is.na(time_median), .(op, n, trees, ref,
      metric = "time (s)", value = time_median)]
  )
  for (o in sort(unique(long$op))) {
    cat("\n\n### ", o, "()\n\n", sep = "")
    dd <- long[op == o]
    ss <- seq_long[op == o]
    p <- ggplot(dd, aes(workers, value, colour = backend, shape = ref)) +
      facet_wrap(~ metric + n + trees, scales = "free", labeller = label_both) +
      geom_line(aes(group = interaction(backend, ref)), linetype = "dotted", linewidth = 0.5) +
      geom_point(size = 2.6) +
      scale_x_continuous(breaks = sort(unique(dd$workers))) +
      scale_colour_brewer(palette = "Dark2") +
      labs(x = "workers", y = NULL, title = paste0(o, "(): backends by worker count"),
        caption = "Dotted lines group a series; they do not interpolate. Dashed grey: the sequential run.") +
      bench_theme()
    if (nrow(ss)) {
      p <- p + geom_hline(data = ss, aes(yintercept = value),
        linetype = "dashed", colour = "grey40")
    }
    print(p)
    cat("\n\n")
  }
}
```

## Per-cell measurements

```{r}
#| label: latest-plots
#| fig-height: 8
if (have_run) {
  long <- rbind(
    run[!is.na(peak_delta_mb), .(op, backend, workers, n, trees, ref,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    run[!is.na(time_median), .(op, backend, workers, n, trees, ref,
      metric = "time (s)", value = time_median)]
  )
  long[, cell := paste0(backend, fifelse(is.na(workers), "", paste0(" w", workers)))]
  print(
    ggplot(long, aes(interaction(n, trees), value, colour = ref, group = ref)) +
      geom_point(size = 2.4) +
      geom_line(linetype = "dotted", linewidth = 0.5) +
      facet_wrap(~ metric + op, scales = "free", labeller = label_both) +
      scale_colour_brewer(palette = "Dark2") +
      labs(x = "n . trees", y = NULL, title = "Per-cell measurements by ref") +
      bench_theme() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

## Can these cells resolve memory?

A memory delta only means something when the operation's own allocation is
large next to the fixed floor. Measured same-commit spread against that ratio:
0.29 gave 14-23%, 0.73 gave 4.2%, and 2.0 gave at most 3.6%. Cells below the
dashed line get no memory verdict.

```{r}
#| label: resolution
#| fig-height: 8
if (is.null(run)) {
  cat("No run loaded.\n")
} else {
  rr <- run[!is.na(peak_delta_mb) & !is.na(floor_mb) & floor_mb > 0]
  if (!nrow(rr)) {
    cat("No cells with both a marginal and a floor.\n")
  } else {
    rr[, ratio := peak_delta_mb / floor_mb]
    rr[, cell := paste0(op, " ", backend, fifelse(is.na(workers), "", paste0(" w", workers)),
      " n=", n, " t=", trees)]
    print(
      ggplot(rr, aes(reorder(cell, ratio), ratio, colour = ref)) +
        geom_hline(yintercept = BENCH_MEM_RESOLUTION, linetype = "dashed") +
        geom_point(size = 2.4) +
        coord_flip() +
        scale_y_log10() +
        scale_colour_brewer(palette = "Dark2") +
        labs(x = NULL, y = "marginal / floor (log)", title = "Memory resolution per cell") +
        bench_theme()
    )
  }
}
```

```{r}
#| label: resolution-table
if (!is.null(run)) {
  rr <- run[!is.na(peak_delta_mb) & !is.na(floor_mb) & floor_mb > 0]
  if (nrow(rr)) {
    rr[, ratio := peak_delta_mb / floor_mb]
    bench_tbl(rr[, .(
      ratio = round(median(ratio), 2),
      resolvable = median(ratio) >= BENCH_MEM_RESOLUTION
    ), by = .(op, n)][order(op, n)])
  }
}
```

## Release trend

Anchors are re-measured on one node, so this is not stitched across machines.
Each box covers every cell measured for that version and size; points are the
cells themselves.

```{r}
#| label: history-trend
#| fig-height: 8
hist <- if (file.exists("history.csv")) as.data.table(fread("history.csv")) else data.table()
if (!nrow(hist)) {
  cat("No anchor history yet; run the suite and commit `bench/history.csv`.\n")
} else {
  long <- rbind(
    hist[!is.na(peak_delta_mb), .(arf_version, op, n, host,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    hist[!is.na(time_median), .(arf_version, op, n, host,
      metric = "time (s)", value = time_median)]
  )
  long[, size := factor(n)]
  print(
    ggplot(long, aes(arf_version, value, colour = size, fill = size)) +
      geom_boxplot(alpha = 0.35, outlier.shape = NA, position = position_dodge(width = 0.8)) +
      geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.15),
        size = 1.8, alpha = 0.9) +
      facet_wrap(~ metric + op, scales = "free", labeller = label_both) +
      scale_y_log10() +
      scale_colour_brewer(palette = "Dark2") +
      scale_fill_brewer(palette = "Dark2") +
      labs(x = NULL, y = NULL, title = "arf performance across anchors",
        caption = "Box over the cells measured for that version and size; points are the cells.") +
      bench_theme() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

```{r}
#| label: history-table
if (nrow(hist)) {
  bench_tbl(hist[, .(
    cells = .N, hosts = uniqueN(host),
    metric = paste(unique(metric), collapse = ", ")
  ), by = .(arf_version, op)][order(arf_version, op)])
}
```
