---
title: "ARF backend benchmarks"
subtitle: "Runtime and peak memory across backends and worker counts"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 8
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
suppressMessages({
  library(data.table)
  library(ggplot2)
})

# Reuse viz.R's file selection: newest sweep-ops CSV per config label, or files
# passed via the QUARTO_VIZ_FILES env var (space-separated) for ad-hoc renders.
files <- Sys.getenv("QUARTO_VIZ_FILES")
files <- if (nzchar(files)) {
  strsplit(files, " ")[[1]]
} else {
  f <- list.files(
    "results",
    pattern = "^sweep-ops-.*\\.csv$",
    full.names = TRUE
  )
  if (!length(f)) {
    stop("no sweep-ops CSVs in bench/results/")
  }
  info <- data.table(file = f)
  info[, base := sub("\\.csv$", "", sub("^sweep-ops-?", "", basename(file)))]
  info[, stamp := sub("^.*?([0-9]{8}-[0-9]{6})$", "\\1", base)]
  info[, label := sub("-?[0-9]{8}-[0-9]{6}$", "", base)]
  info[order(-stamp), .SD[1], by = label]$file
}

res <- rbindlist(lapply(files, fread), fill = TRUE)
if (!"op" %in% names(res)) {
  res[, op := NA_character_]
}
res[is.na(op), op := "forde"] # old forde-only sweep.R CSVs
res[, minutes := seconds / 60]
res[, gib := peak_mb / 1024]

# forge/expct differ by rowmode and evidence/synth scale -> a "variant" label so
# those become nested tabs within the op instead of extra facets.
res[, variant := NA_character_]
if ("rowmode" %in% names(res)) {
  res[
    op %in% c("forge", "expct") & !is.na(rowmode),
    variant := paste0(
      rowmode,
      " (evi=",
      n_evidence,
      fifelse(!is.na(n_synth), paste0(", synth=", n_synth), ""),
      ")"
    )
  ]
}

seq_res <- res[backend == "sequential"] # one measurement per (op, variant, n, trees)
par_res <- res[backend != "sequential"] # workers series

# One leaf plot: a single op (+ variant), faceted trees x n, workers on x.
# y is always log10 (values span orders of magnitude across n and trees).
plot_leaf <- function(op_val, variant_val, yvar, ylab) {
  keep <- function(dt) {
    dt[
      op == op_val &
        (is.na(variant_val) | variant == variant_val)
    ]
  }
  pr <- keep(par_res)
  sq <- keep(seq_res)
  p <- ggplot(
    pr,
    aes(x = workers, y = .data[[yvar]], color = backend, fill = backend)
  ) +
    facet_grid(trees ~ n, labeller = label_both, scales = "free_y") +
    geom_hline(
      data = sq,
      aes(yintercept = .data[[yvar]]),
      color = "blue",
      linetype = "dashed"
    ) +
    geom_line() +
    geom_point(size = 2.5, key_glyph = "rect") +
    scale_x_continuous(breaks = sort(unique(pr$workers))) +
    labs(
      x = "# of workers",
      y = ylab,
      color = NULL,
      fill = NULL,
      caption = "Dashed blue: sequential run"
    ) +
    scale_y_log10() +
    theme_minimal(base_size = 16) +
    theme(legend.position = "bottom", plot.title.position = "plot")
  p
}

metrics <- list(
  list(yvar = "minutes", ylab = "Time (minutes, log10)", title = "Runtime"),
  list(yvar = "gib", ylab = "Peak memory (GiB, log10)", title = "Peak memory")
)
ops <- sort(unique(res$op))
```

One section per arf function. Within it, tabs switch between the runtime and
peak-memory views (and, for `forge`/`expct`, the evidence/synth variant) so all
plots for a function sit side by side. Each panel is faceted by number of trees
(rows) and sample size `n` (columns); coloured lines are the parallel backends
across worker counts, and the dashed blue line marks the sequential baseline.

```{r}
#| label: tabsets
#| output: asis
# Function-first: op is a real section (TOC-jumpable); its plots are fenced
# `::: {.panel-tabset}` divs (the only form Quarto turns into interactive tabs
# from an asis chunk). Outer fence needs more colons than the nested one.
for (op_val in ops) {
  cat("\n\n##", paste0(op_val, "()"), "\n\n")
  variants <- unique(res[op == op_val]$variant)
  single <- length(variants) == 1 && all(is.na(variants))
  cat("\n::::: {.panel-tabset}\n")
  for (m in metrics) {
    cat("\n###", m$title, "\n\n")
    if (single) {
      print(plot_leaf(op_val, variants[[1]], m$yvar, m$ylab))
      cat("\n\n")
    } else {
      cat("\n:::: {.panel-tabset}\n")
      for (v in variants) {
        cat("\n####", if (is.na(v)) "default" else v, "\n\n")
        print(plot_leaf(op_val, v, m$yvar, m$ylab))
        cat("\n\n")
      }
      cat("\n::::\n")
    }
  }
  cat("\n:::::\n")
}
```
