---
title: "arf benchmarks"
subtitle: "Version comparison, backend comparison, and the release trend"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    embed-resources: true
    fig-width: 11
    fig-height: 7
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
suppressMessages({
  library(data.table)
  library(ggplot2)
})
source("run-cell.R")
source("collate.R")

# Small tables render as plain HTML; big ones get a sortable, filterable widget
# rather than a wall of rows.
bench_tbl <- function(x, big = 25L, page = 15L) {
  x <- as.data.frame(x)
  if (!nrow(x)) {
    return(htmltools::HTML("<p><em>No rows.</em></p>"))
  }
  if (nrow(x) >= big && requireNamespace("reactable", quietly = TRUE)) {
    reactable::reactable(
      x,
      filterable = TRUE, sortable = TRUE, searchable = TRUE,
      defaultPageSize = page, compact = TRUE, striped = TRUE, highlight = TRUE,
      defaultColDef = reactable::colDef(minWidth = 70)
    )
  } else {
    htmltools::HTML(knitr::kable(x, format = "html", table.attr = 'class="table table-sm"'))
  }
}

bench_theme <- function(base_size = 15) {
  theme_minimal(base_size = base_size) +
    theme(
      legend.position = "bottom",
      legend.title = element_blank(),
      plot.title.position = "plot",
      # Facet labels on the left rather than the right, where they read as
      # detached from the panel they describe.
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0),
      panel.spacing = unit(1, "lines")
    )
}

# Newest version-comparison CSV, or one named via QUARTO_BENCH_CSV.
bench_csv <- Sys.getenv("QUARTO_BENCH_CSV")
if (!nzchar(bench_csv)) {
  cands <- list.files("results", pattern = "^bench-.*\\.csv$", full.names = TRUE)
  bench_csv <- if (length(cands)) cands[order(basename(cands), decreasing = TRUE)][1] else ""
}
run <- if (nzchar(bench_csv) && file.exists(bench_csv)) as.data.table(fread(bench_csv)) else NULL
baseline <- Sys.getenv("QUARTO_BENCH_BASELINE", "main")
have_run <- !is.null(run) && baseline %in% run$ref
```

::: {.callout-note collapse="true"}
## What the memory numbers mean

Every cell is measured twice per round: once running the operation, and once
running nothing but the setup (an R interpreter, `arf`, and the fixture data).
That second measurement is the **floor**, about 200 MB here. The **marginal**
is peak minus floor: what the operation itself allocated, as opposed to the
fixed cost of having R and the package loaded.

Deltas are computed on the marginal, because the floor is identical across refs
and would dilute any real difference several-fold. A marginal smaller than its
own floor is below what this instrument can resolve, and gets no verdict.
:::

## Latest version comparison

```{r}
#| label: latest-run
if (is.null(run)) {
  cat("No `results/bench-*.csv` yet. Run `bench/arf-bench run`.\n")
} else if (!have_run) {
  cat("Baseline `", baseline, "` is not in ", basename(bench_csv),
    "; set `--base` or QUARTO_BENCH_BASELINE.\n", sep = "")
} else {
  cat(sprintf(
    "`%s` | %d rows | refs: %s | metric: %s | host(s): %s\n\n",
    basename(bench_csv), nrow(run), paste(unique(run$ref), collapse = ", "),
    paste(unique(run$metric), collapse = ", "),
    paste(unique(run$host), collapse = ", ")
  ))
}
```

```{r}
#| label: latest-table
if (have_run) {
  d <- as.data.table(bench_deltas(run, baseline = baseline))[ref != baseline]
  bench_tbl(d[order(op, n, trees, backend, workers, ref), .(
    op, backend, w = workers, n, trees, ref,
    marginal_MB = round(peak_delta_mb, 1),
    floor_MB = round(floor_mb, 1),
    mem_pct = round(delta_mem_pct, 1),
    mem = mem_verdict,
    time_s = round(time_median, 3),
    time_pct = round(delta_time_pct, 1),
    time = time_verdict
  )])
}
```

## Backend comparison

Which backend, at which worker count. This is a live question, not a historical
one: `foreach` forks and copies, `mirai` shares read-only through `mori`, and
which wins depends on data size and worker count. Worker counts are discrete
measurements, so the connecting lines are dotted: they group a series, they do
not interpolate.

```{r}
#| label: backend-compare
#| output: asis
#| fig-height: 9
if (!have_run) {
  cat("No run loaded.\n")
} else if (uniqueN(run$backend) < 2L) {
  cat("This run used a single backend (`", unique(run$backend),
    "`), so there is nothing to compare. The full tier varies backend and workers.\n", sep = "")
} else {
  par_run <- run[backend != "sequential" & !is.na(workers)]
  seq_run <- run[backend == "sequential"]
  long <- rbind(
    par_run[!is.na(peak_delta_mb), .(op, n, trees, backend, workers, ref,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    par_run[!is.na(time_median), .(op, n, trees, backend, workers, ref,
      metric = "time (s)", value = time_median)]
  )
  seq_long <- rbind(
    seq_run[!is.na(peak_delta_mb), .(op, n, trees, ref,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    seq_run[!is.na(time_median), .(op, n, trees, ref,
      metric = "time (s)", value = time_median)]
  )
  for (o in sort(unique(long$op))) {
    cat("\n\n### ", o, "()\n\n", sep = "")
    dd <- long[op == o]
    ss <- seq_long[op == o]
    p <- ggplot(dd, aes(workers, value, colour = backend, shape = ref)) +
      facet_wrap(~ metric + n + trees, scales = "free", labeller = label_both) +
      geom_line(aes(group = interaction(backend, ref)), linetype = "dotted", linewidth = 0.5) +
      geom_point(size = 2.6) +
      scale_x_continuous(breaks = sort(unique(dd$workers))) +
      scale_colour_brewer(palette = "Dark2") +
      labs(x = "workers", y = NULL, title = paste0(o, "(): backends by worker count"),
        caption = "Dotted lines group a series; they do not interpolate. Dashed grey: the sequential run.") +
      bench_theme()
    if (nrow(ss)) {
      p <- p + geom_hline(data = ss, aes(yintercept = value),
        linetype = "dashed", colour = "grey40")
    }
    print(p)
    cat("\n\n")
  }
}
```

## Per-cell measurements

```{r}
#| label: latest-plots
#| fig-height: 8
if (have_run) {
  long <- rbind(
    run[!is.na(peak_delta_mb), .(op, backend, workers, n, trees, ref,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    run[!is.na(time_median), .(op, backend, workers, n, trees, ref,
      metric = "time (s)", value = time_median)]
  )
  long[, cell := paste0(backend, fifelse(is.na(workers), "", paste0(" w", workers)))]
  print(
    ggplot(long, aes(interaction(n, trees), value, colour = ref, group = ref)) +
      geom_point(size = 2.4) +
      geom_line(linetype = "dotted", linewidth = 0.5) +
      facet_wrap(~ metric + op, scales = "free", labeller = label_both) +
      scale_colour_brewer(palette = "Dark2") +
      labs(x = "n . trees", y = NULL, title = "Per-cell measurements by ref") +
      bench_theme() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

## Can these cells resolve memory?

A memory delta only means something when the operation's own allocation is
large next to the fixed floor. Measured same-commit spread against that ratio:
0.29 gave 14-23%, 0.73 gave 4.2%, and 2.0 gave at most 3.6%. Cells below the
dashed line get no memory verdict.

```{r}
#| label: resolution
#| fig-height: 8
if (is.null(run)) {
  cat("No run loaded.\n")
} else {
  rr <- run[!is.na(peak_delta_mb) & !is.na(floor_mb) & floor_mb > 0]
  if (!nrow(rr)) {
    cat("No cells with both a marginal and a floor.\n")
  } else {
    rr[, ratio := peak_delta_mb / floor_mb]
    rr[, cell := paste0(op, " ", backend, fifelse(is.na(workers), "", paste0(" w", workers)),
      " n=", n, " t=", trees)]
    print(
      ggplot(rr, aes(reorder(cell, ratio), ratio, colour = ref)) +
        geom_hline(yintercept = BENCH_MEM_RESOLUTION, linetype = "dashed") +
        geom_point(size = 2.4) +
        coord_flip() +
        scale_y_log10() +
        scale_colour_brewer(palette = "Dark2") +
        labs(x = NULL, y = "marginal / floor (log)", title = "Memory resolution per cell") +
        bench_theme()
    )
  }
}
```

```{r}
#| label: resolution-table
if (!is.null(run)) {
  rr <- run[!is.na(peak_delta_mb) & !is.na(floor_mb) & floor_mb > 0]
  if (nrow(rr)) {
    rr[, ratio := peak_delta_mb / floor_mb]
    bench_tbl(rr[, .(
      ratio = round(median(ratio), 2),
      resolvable = median(ratio) >= BENCH_MEM_RESOLUTION
    ), by = .(op, n)][order(op, n)])
  }
}
```

## Release trend

Anchors are re-measured on one node, so this is not stitched across machines.
Each box covers every cell measured for that version and size; points are the
cells themselves.

```{r}
#| label: history-trend
#| fig-height: 8
hist <- if (file.exists("history.csv")) as.data.table(fread("history.csv")) else data.table()
if (!nrow(hist)) {
  cat("No anchor history yet; run the suite and commit `bench/history.csv`.\n")
} else {
  long <- rbind(
    hist[!is.na(peak_delta_mb), .(arf_version, op, n, host,
      metric = "marginal memory (MB)", value = peak_delta_mb)],
    hist[!is.na(time_median), .(arf_version, op, n, host,
      metric = "time (s)", value = time_median)]
  )
  long[, size := factor(n)]
  print(
    ggplot(long, aes(arf_version, value, colour = size, fill = size)) +
      geom_boxplot(alpha = 0.35, outlier.shape = NA, position = position_dodge(width = 0.8)) +
      geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.15),
        size = 1.8, alpha = 0.9) +
      facet_wrap(~ metric + op, scales = "free", labeller = label_both) +
      scale_y_log10() +
      scale_colour_brewer(palette = "Dark2") +
      scale_fill_brewer(palette = "Dark2") +
      labs(x = NULL, y = NULL, title = "arf performance across anchors",
        caption = "Box over the cells measured for that version and size; points are the cells.") +
      bench_theme() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

```{r}
#| label: history-table
if (nrow(hist)) {
  bench_tbl(hist[, .(
    cells = .N, hosts = uniqueN(host),
    metric = paste(unique(metric), collapse = ", ")
  ), by = .(arf_version, op)][order(arf_version, op)])
}
```

## Earlier backend sweeps (`sweep-ops-*.csv`)

Runs from the standalone sweep harness (`sweep-ops.R`), which measures backends
within a single version rather than across refs. The backend question it answers
is the same live one as the section above; these files exist because those
sweeps covered sizes and worker counts the regression grid does not. Skipped
when no such CSVs are present.

```{r}
#| label: legacy-setup
legacy_files <- Sys.getenv("QUARTO_VIZ_FILES")
legacy_files <- if (nzchar(legacy_files)) {
  strsplit(legacy_files, " ")[[1]]
} else {
  f <- list.files("results", pattern = "^sweep-ops-.*\\.csv$", full.names = TRUE)
  if (!length(f)) {
    character()
  } else {
    info <- data.table(file = f)
    info[, base := sub("\\.csv$", "", sub("^sweep-ops-?", "", basename(file)))]
    info[, stamp := sub("^.*?([0-9]{8}-[0-9]{6})$", "\\1", base)]
    info[, label := sub("-?[0-9]{8}-[0-9]{6}$", "", base)]
    info[order(-stamp), .SD[1], by = label]$file
  }
}
have_sweeps <- length(legacy_files) > 0L
if (!have_sweeps) {
  cat("No `sweep-ops-*.csv` in results/.\n")
  # Everything below this point belongs to the legacy section, so switching
  # evaluation off here skips it without touching the sections above.
  knitr::opts_chunk$set(eval = FALSE)
} else {
  res <- rbindlist(lapply(legacy_files, fread), fill = TRUE)
  # Guard against apples/oranges: warn when the loaded CSVs mix memory metrics
  # (PSS vs cgroup-anon) or arf code states (commit column, incl. +dirty).
  for (col in c("metric", "commit")) {
    if (col %in% names(res) && uniqueN(res[[col]], na.rm = TRUE) > 1) {
      warning("mixing ", col, "s across CSVs: ", paste(unique(na.omit(res[[col]])), collapse = ", "), call. = FALSE)
    }
  }
  if (!"op" %in% names(res)) {
    res[, op := NA_character_]
  }
  res[is.na(op), op := "forde"] # old forde-only sweep.R CSVs
  res[, minutes := seconds / 60]
  res[, gib := peak_mb / 1024]

  # forge/expct differ by rowmode and evidence/synth scale -> a "variant" label so
  # those become nested tabs within the op instead of extra facets.
  res[, variant := NA_character_]
  if ("rowmode" %in% names(res)) {
    res[
      op %in% c("forge", "expct") & !is.na(rowmode),
      variant := paste0(
        rowmode,
        " (evi=",
        n_evidence,
        fifelse(!is.na(n_synth), paste0(", synth=", n_synth), ""),
        ")"
      )
    ]
  }

  seq_res <- res[backend == "sequential"] # one measurement per (op, variant, n, trees)
  par_res <- res[backend != "sequential"] # workers series

  # One leaf plot: a single op (+ variant), faceted trees x n, workers on x.
  # y is always log10 (values span orders of magnitude across n and trees).
  plot_leaf <- function(op_val, variant_val, yvar, ylab) {
    keep <- function(dt) {
      dt[
        op == op_val &
          (is.na(variant_val) | variant == variant_val)
      ]
    }
    pr <- keep(par_res)
    sq <- keep(seq_res)
    p <- ggplot(
      pr,
      aes(x = workers, y = .data[[yvar]], color = backend, fill = backend)
    ) +
      # facet_grid frees y per ROW only, so panels in a row shared a scale
      # and squashed; facet_wrap frees every panel independently.
      facet_wrap(~ trees + n, labeller = label_both, scales = "free") +
      geom_hline(
        data = sq,
        aes(yintercept = .data[[yvar]]),
        color = "blue",
        linetype = "dashed"
      ) +
      # Dotted: worker counts are separate measurements, not a continuum, and
      # a solid line reads as a trend (e.g. "foreach falls with workers")
      # where the movement may be noise.
      geom_line(linetype = "dotted", linewidth = 0.5) +
      geom_point(size = 2.5, key_glyph = "rect") +
      scale_x_continuous(breaks = sort(unique(pr$workers))) +
      labs(
        x = "# of workers",
        y = ylab,
        color = NULL,
        fill = NULL,
        caption = "Dashed blue: sequential run"
      ) +
      scale_y_log10() +
      scale_colour_brewer(palette = "Dark2") +
      scale_fill_brewer(palette = "Dark2") +
      bench_theme()
    p
  }

  metrics <- list(
    list(yvar = "minutes", ylab = "Time (minutes, log10)", title = "Runtime"),
    list(yvar = "gib", ylab = "Peak memory (GiB, log10)", title = "Peak memory")
  )
  ops <- sort(unique(res$op))
}
```

One section per arf function. Within it, tabs switch between the runtime and
peak-memory views (and, for `forge`/`expct`, the evidence/synth variant) so all
plots for a function sit side by side. Each panel is faceted by number of trees
(rows) and sample size `n` (columns); coloured lines are the parallel backends
across worker counts, and the dashed blue line marks the sequential baseline.

```{r}
#| label: tabsets
#| output: asis
# Function-first: op is a real section (TOC-jumpable); its plots are fenced
# `::: {.panel-tabset}` divs (the only form Quarto turns into interactive tabs
# from an asis chunk). Outer fence needs more colons than the nested one.
for (op_val in ops) {
  cat("\n\n##", paste0(op_val, "()"), "\n\n")
  variants <- unique(res[op == op_val]$variant)
  single <- length(variants) == 1 && all(is.na(variants))
  cat("\n::::: {.panel-tabset}\n")
  for (m in metrics) {
    cat("\n###", m$title, "\n\n")
    if (single) {
      print(plot_leaf(op_val, variants[[1]], m$yvar, m$ylab))
      cat("\n\n")
    } else {
      cat("\n:::: {.panel-tabset}\n")
      for (v in variants) {
        cat("\n####", if (is.na(v)) "default" else v, "\n\n")
        print(plot_leaf(op_val, v, m$yvar, m$ylab))
        cat("\n\n")
      }
      cat("\n::::\n")
    }
  }
  cat("\n:::::\n")
}
```
